---
title: "GSLS - CS2 Exercises"
author: "Liana Hardy"
date: '2020-05-08'
output:
  pdf_document: default
  html_notebook: default
---

## 2.9 Exercise
Juvenile lobsters in aquaculture were grown on three different diets (fresh mussels, semi-dry pellets and dry flakes). After nine weeks, their wet weight was:

```{r}
lobstersdf<-read.csv(file='CS2-lobsters.csv')
lobstersdf
```

### Hypotheses

* H0: The mean weight of lobsters of all diets is the same μM = μP = μF.
* H1: The mean weights are not all equal.

### Aggregate, Summarise and Visualise
```{r}
aggregate(weight~diet, data=lobstersdf, summary)
summary(lobstersdf)
boxplot(weight~diet, data=lobstersdf)
```
### Perform an ANOVA on linearised data
```{r}
lm.lobster<-lm(weight~diet, data=lobstersdf)
anova(lm.lobster)
```
__A one-way ANOVA showed that the mean weight of lobsters is not significantly different between diets (F=1.64, df=2,15, p=0.2263)__

### Check Assumptions - normality  

```{r}
par(mfrow=c(2,2))
plot(lm.lobster)
```


#### Unstack the data
```{r}
uns.lobsters<-unstack(lobstersdf)
uns.lobsters
```

#### Perform Shapiro-wilks on each data set
```{r}
shapiro.test(uns.lobsters$Flakes)
shapiro.test(uns.lobsters$Mussels)
shapiro.test(uns.lobsters$Pellets)

qqnorm(uns.lobsters$Pellets)
qqline(uns.lobsters$Pellets)
```
__Shapiro.wilks test shows Flakes and Mussels datasets show normal distribution, however Pellets <0.05. Shapiro-Wilks is sensitive to sample size. Q-Q plot shows an outlier but overall with p-value = 0.0425 and Q-Q plot I am happy with the assumption that these data-sets are normally distributed.__


### Check assumptions - variance

```{r}
bartlett.test(weight~diet, data=lobstersdf)
```
__Bartletts test of homogeneity of variances infers that data sets have equal variances.__

### Perform Post-Hoc test

#### Tukeys test
```{r}
aov.lobster<-aov(weight~diet, data=lobstersdf)
TukeyHSD(aov.lobster)
```

### Report the findings 

__One-way ANOVA and Tukeys test shows that the mean weight of lobsters is not different between any of the three diets (p>0.05).__

*****

### Perform Kruskal-Wallis test
```{r}
kruskal.test(weight~diet, data=lobstersdf)
```

#### Perform dunn.test - post-hoc test
```{r}
library(dunn.test)
dunn.test(lobstersdf$weight, lobstersdf$diet)
```

__Kruskal-Wallis and dunn test show that weight of lobsters fed on mussels is significantly different than flakes (p-value = 0.037)__

## Revision Exercise

#### Open clean data csv file 

```{r}
drugdf<-read.csv(file="cs2revision_data_clean.csv",header=T)
drugdf
```
#### Aggregate, visualise and summarise the data
```{r}
aggregate(Value~Drug, data=drugdf, summary)
summary(drugdf)
boxplot(Value~Drug, data=drugdf)
```
#### Perform ANOVA
```{r}
lm.drugs<-lm(Value~Drug, data=drugdf)
anova(lm.drugs)
```

#### Diagnostic plots
```{r}
par(mfrow=c(2,2))
plot(lm.drugs)
```


```{r}
aov.drugs<-aov(Value~Drug, data=drugdf)
TukeyHSD(aov.drugs)

```
__ Assumptions for ANOVA and Tukeys test are met so we can say that these drugs do not differ significantly__
