---
title: "GSLS - CS1 Exercises"
author: "Liana Hardy"
date: '2020-05-02'
output: pdf_document
---

## 3.3.7 Exercise on One-Sample t-test

The following data are the dissolving times (in seconds) of a drug in agitated gastric juice:
42.7, 43.4, 44.6, 45.1, 45.6, 45.9, 46.8, 47.6

Do these results provide any evidence to suggest that dissolving time for this drug is different from 45s?

1. Write down the null and alternative hypotheses.

H0: The mean dissolving time is equal to 45s ((μ=45)
H1: The mean body length is not equal to 45s (μ≠45).

2. Summarise and visualise the data and perform an appropriate one-sample t-test.

3. What can you say about the dissolving time? (what sentence would you use to report this)

A one-sample t-test indicated that the mean dissolving time of drug in an agitated stomach (μ=45.2125) is not significantly different from 45 (t=0.36647, df=7, p=0.7248).

4. Check the assumptions for the test.
- histogram
- qq plot

Graphs suggest normal distribution

5. Was the test valid?

- shapiro-wilks
As the p-value is bigger than 0.05 - there is insufficient evidence to reject the null hypothesis that the sample came from a normal distribution.

## Create a vector and name it dissolving_time
```{r}
dissolving_time <- c(42.7, 43.4, 44.6, 45.1, 45.6, 45.9, 46.8, 47.6) # units = seconds
dissolving_time
```

## Create a dataframe
```{r}
dissolving_timeDF <- data.frame(dissolving_time)
str(dissolving_timeDF)
```

## Summarise and visualise the data
```{r}
summary(dissolving_timeDF)
boxplot(dissolving_timeDF, main="Drug", ylab="Dissolving Time (s)")

```
## Perform a one-sample, two-tailed t-test
```{r}
t.test(dissolving_timeDF, mu=45, alternative="two.sided")

# first argument must be a numerical vector of data values 
# second argument must be a number and is the mean to be tested under the null hypothesis 
# third argument gives the type of alternative hypothesis and must be one of "two.sided","greater" or "less"
```

## Create a histogram to check normality 
```{r}
hist(dissolving_time, breaks = 5)
```
## Create a QQ plot to check normality
```{r}
qqnorm(dissolving_time)
qqline(dissolving_time)
```

## Perform a Shapiro-Wilk Test
```{r}
shapiro.test(dissolving_time)
```
3.4.6 Exercise on One-Sample Wilcoxon signed-rank test

## Analyse the drug dataset from before using a one-sample Wilcoxon signed-rank test?
```{r}
wilcox.test(dissolving_time, mu=45, alternative = "two.sided")
```

## Plot a histogram and boxplot from the data

```{r}
hist(dissolving_time, breaks = 5)
boxplot(dissolving_time)
```


A one-sample Wilcoxon signed-rank test indicated that the median dissolving time of a drug in an agitated stomach (μ=45.35s) does not differ significantly from 45s (V=22, n=8, p=0.6406). 

Discuss with a neighbour which of the two tests you feel is best suited to the data.

QQ plot not very conclusive, some snaking - Shamiro-wilks test has insignificant p-value (p = 0.9641) so agree that assumptions of test is met (normal distribution). Prefer this to wilcoxon signed-rank test as t-test holds more power and assumption of symmetry in boxplots/qq plot was not convincing. 

Does it matter in this case?

Not really both show high p-value/ not significant 

## 4.3.7 Exercise on two-sample t-test

1. Write down the null and alternative hypotheses.

H0: Male and female turtles have the same mean serum cholesterol concentrations
H1: Male and female turtles have different mean serum cholesterol concentrations

Using the following data:

# write data in stacked format and into data frame
```{r}

serumchol <- c(233.4, 221.5, 230.2, 224.3, 223.8, 230.8, 220.1, 218.6, 229.6, 228.8, 222.0, 224.1, 226.5)
gender <- c("Female","Female","Female","Female","Female","Female", "Male","Male","Male","Male","Male","Male","Male")
data.frame(serumchol, gender) -> turtledf
turtledf
```
## visualise and summarise the data
```{r}
aggregate(serumchol~gender, data=turtledf, summary)
boxplot(serumchol~gender, data=turtledf,  ylab= "Concentration (mg/100ml)", xlab = "Gender", main="Serum Cholesterol")
        
```

## Perform a two-sample, two-tailed t-test
```{r}
t.test(serumchol~gender, data=turtledf, alternative="two.sided", var.equal = TRUE)
```
## Unstack and repeat t-test
```{r}
unstack(turtledf)-> uns.turtle
uns.turtle
t.test(uns.turtle$F, uns.turtle$M, alternative="two.sided", var.equal= TRUE)
```
ANSWER: Students t-test indicates that the mean serum cholesterol of male (224.2) and female turtles (227.3) is not significantly different (t=1.2383, df=11, p-value = 0.2414).

## check normality - shapiro.wilks test
```{r}
shapiro.test(uns.turtle$Female)
shapiro.test(uns.turtle$Male)
```
ANSWER: both male and female data sets are normally distributed.

## check variance - Bartlett.test
```{r}
bartlett.test(serumchol~gender, data=turtledf)
```
ANSWER: Probability of observing these two samples if they come from distributions with the same variance (>0.05). Students t-test is valid. If this was not the case, a welch's t-test would have been performed. 

## Perform Mann-Whitney U test
```{r}
wilcox.test(serumchol~gender, data=turtledf, alternative="two.sided")
```

ANSWER: not significant

## Check assumptions are correct to use a paired t-test for cortisol data

## Normality - shapiro
```{r}
cortisol
shapiro.test(cortisol$morning)
shapiro.test(cortisol$evening)
```
ANSWER: Morning and evening cortisol levels have normal distribution (>0.05), use bartlett test for variance


## Variance - bartlett
```{r}
bartlett.test(cortisol)
```

p-value >0.05, variances are the same cannot dismiss null. Paired t-test was valid. 

## 4.6.5 Exercise - paired data

H0: There is no difference in leg length between hind and fore legs (μHL = μFL).
H1: There is a difference in leg length between hind and fore legs (μHL ≠ μFL).

## Write deer data into R

```{r}
hindleg <- c(142,140,144,144,142,146,149,150,142,148) # length in cm
foreleg <-c(138,136,147,139,143,141,143,145,136,146) # length in cm
pairdeerleg <- data.frame(hindleg, foreleg)

```

## Summarise and visualise the data
```{r}
summary(pairdeerleg)
boxplot(pairdeerleg)
changedeerleg <- pairdeerleg$hindleg - pairdeerleg$foreleg
boxplot(changedeerleg, ylab = "Change in leg length (cm)", main = "Difference in deer fore and hind legs")
```
## Perform a two-sample paired t-test - assume normal distribution

```{r}
t.test(pairdeerleg$hindleg, pairdeerleg$foreleg, alternative="two.sided", paired=T)
```

## Perform wilcoxon signed rank test
```{r}
wilcox.test(pairdeerleg$hindleg, pairdeerleg$foreleg, alternative = "two.sided", paired=T)
```

## Check assumptions - normality
```{r}
qqnorm(pairdeerleg$hindleg)
qqline(pairdeerleg$hindleg)
shapiro.test(pairdeerleg$hindleg)

```
```{r}
qqnorm(pairdeerleg$foreleg)
qqline(pairdeerleg$foreleg)
shapiro.test(pairdeerleg$foreleg)
```
ANSWER: p-value > 0.05 data is normally distributed. 

## Variance - bartlett test (normal distribution)
```{r}
bartlett.test(pairdeerleg)
```
ANSWER: p-value >0.05 suggests equal variances in paired samples. 

## Discuss which test is most appropriate

1. Are the samples normally distributed? (Yes)
2. How big are your samples? (<30 data points)

Small data set with normal distribution - use parametric test

3. Are the samples paired? (Yes)
4. Do the samples have the same variance? (Yes)

Samples are paired and have the same variance therefore paired t-test is most appropriate. 

## Summarise the results.

A two-tailed, paired t-test indicates that the mean leg length of deers differed significantly between the hindleg (144.7cm) and foreleg (141.4cm) (t = 3.4138, df = 9, p-value = 0.007703).
